\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-07-29 NAR_DRAFT.tex
found a citation for different scales of genomic context

\subsection{Estimation of the mutation profile}
There are many aspects of local genomic architecture, generally described as the genomic \textit{context}, which can affect the local somatic mutation rate at a particular site in the genome.
Here we focus on the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, and three possible alternate alleles {\{A,(C/T),G\}} for each.
Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles.

@@

\subsection{Estimation of the mutation rate.}
The final piece of equation \ref{eqn:full_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth and death rate.
\citet{Williams2016} showed that if 
$$
  \beta = \frac{b-d}{b},
$$
then the number of new mutations per cell division is
$$
  \frac{N \mu}{\beta}.
$$
In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
$$
  \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
$$
\end{equation}
\bkmcomment{Gets weaker here in terms of how this needs to be said and how much of it needs to be said. Need to decide where to add the factor of two.}

@@

\subsection{Tumor simulations.}

We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from TCGA and PCAWG variant files(dates).
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:

@@

The heart of \batcave, the context-dependent prior probability of mutation, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure \ref{NAR-sigfig} \textbf{e)} and \textbf{f)} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Signature & $\mu/\beta$ & AUROC &  & AUPRC & & ICI & 
\\
& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
\\
\colrule
100X whole genome & Concentrated & 3.6e-7 & Row 1 & -- & -- & -- & -- & --
\\
100X whole genome & Intermediate & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
\\
100X whole genome & Diffuse & Row 1 & Row 1 & -- & -- & -- & -- & --
\\
500X whole exome & Concentrated & Row 1 & .848 & .928 & .674 & .757 & .0585 & .0465
\\
500X whole exome & Intermediate & Row 1 & .847 & .842 & .677 & .707 & .0587 & .0566 
\\
500X whole exome & Diffuse & Row 1 & .850 & .873 & .676 & .698 & .0565 & .0575
\\
AML 31 & Actual & 3.6e-8 & .935 & .949 & .995 & .996 & 0.121 & 0.155
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}

\subsection{Convergence of the mutation prior to the true mutation profile}

The dirichlet prior probability of mutation converges quickly to the true mutation signature.
The rate and degree of convergence are correlated with the concentration of the mutation signature operating in the tumor (Figure \ref{NAR-sigfig}\textbf{e,f}).

@@

>>>>

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \citet{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, and three possible alternate alleles {\{A,(C/T),G\}} for each.
Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles.

@@

\subsection{Estimation of the mutation rate.}
The final piece of equation \ref{eqn:full_prior} to specify is $p(M)$, the prior probability that a mutation will occur at a given site, which usually specified as the mutation \textit{rate} per base $\mu$.
In an exponentially growing tumor, the number of new mutations per cell division is the product of the number of bases $N$, the per base mutation rate $\mu$, and the \textit{effective} birth rate $\beta$, which is a function of the birth $b$ and death $d$ rates.
\citet{Williams2016} showed that if 
$$
  \beta = \frac{b-d}{b},
$$
then the number of new mutations per cell division is
$$
  \frac{N \mu}{\beta}.
$$
In a neutrally or nearly neutrally evolving tumor, they further show that the total number of mutations $\mathrm{M}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
$$
  \mathrm{M}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
$$
\end{equation}
\bkmcomment{Gets weaker here in terms of how this needs to be said and how much of it needs to be said. Need to decide where to add the factor of two.}

@@

\subsection{Tumor simulations.}
\bkmcomment{if $\pi$ works above, describe profile as $\pi$, and $\pi$ as linear combination of 3 signatures.}
We simulated realistic variant sites and allele frequencies using a branching process to simulate neutral evolution with no death.
The simulated mutation rate was $\mu = 3\cdot10^{-6}$.
Variants were selected from TCGA and PCAWG variant files(dates).
Whole genome (100X depth), and whole exome (500X depth) reads were simulated from the GRCH38 reference genome using VarSim \cite{Mu2015}, and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were spiked to create tumors with Bamsurgeon with default parameters \cite{Ewing2015a},
and called with MuTect 1.1.7 \cite{Cibulskis2013} with the following parameters:

@@

The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure \ref{NAR-sigfig} \textbf{e)} and \textbf{f)} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations.

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with $\mu$, auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Mutation profile & $\mu/\beta$ & AUROC & & AUPRC & & ICI & 
\\
& & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
\\
\colrule
100X whole genome & Concentrated & 3.6e-7 & Row 1 & -- & -- & -- & -- & --
\\
100X whole genome & Intermediate & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
\\
100X whole genome & Diffuse & 3.2e-7 & Row 1 & -- & -- & -- & -- & --
\\
500X whole exome & Concentrated & Row 1 & .848 & .928 & .674 & .757 & .0585 & .0465
\\
500X whole exome & Intermediate & Row 1 & .847 & .842 & .677 & .707 & .0587 & .0566 
\\
500X whole exome & Diffuse & Row 1 & .850 & .873 & .676 & .698 & .0565 & .0575
\\
AML 31 & Actual & 3.6e-8 & .935 & .949 & .995 & .996 & 0.121 & 0.155
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}

@@

Simulations suggest that if the mutation profile is the data generating distribution, \batcave can estimate effective mutation rate and data generating distribution, and doing so improves both classification and calibration.
